### 2022-05-04 2022-icfp-unet/Paper/Response.txt

The authors would like to express their gratitude for the reviewers thoughtful responses. We first would like to acknowledge the alien appearance of APL expressions and the challenge that any alien paradigm can present to a reader. We do not wish to claim either that the expressions given in the paper should be considered trivial or obvious to the uninitiated. Much as standard mathematical notation or the use of laziness, static typing, and recursion can present a serious potential challenge to a reader who is unpracticed and unfamiliar with their use, the APL language presents a challenge precisely because it delivers a novel and different approach to solving problems.

We believe, however, that this novelty, and its associated challenges, creates such remarkably unexpected and surprising results that the challenge is worth undertaking. Indeed, the results are remarkable enough that even without taking the time to acquire a complete understanding of the APL expressions included in our paper, the very nature of their construction, the very fact that they can be put into a paper of this size while expressing what they express, makes them worthy of, at the very least, an increased awareness and publication. That is to say, we believe that a reader can derive significant value from our paper simply by becoming aware of what can be done by approaching problems in this manner, even if the exact details remain somewhat unclear without more effort, much in the same way that one could appreciate the potential of algebraic notation after seeing what is possible with the notation even if the only form of reasoning known at that point has been longform prose and considering the associated time it would take to learn mathematical notation.

The reviewers have compared the perceived challenge of our APL expressions against the simpler to call PyTorch API. We contest that this strongly misses the point, and is in fact the wrong comparison. They are comparing the "wiring" code for a neural network, which is no more than the threading of data through pre-defined library functions of specific, limited, and constrained scope, against a fully defined, complete, from scratch, zero-library, zero-framework implementation of that same neural network in a programming language that has no notion of neural networks, no notion of machine learning, and was not designed with machine learning in mind. Our implementation more accurately represents a reimplementation of large portions of the PyTorch functionality. It would be more accurate to imagine what it would be like to reimplement PyTorch's functionality from scratch, with the same performance, and then compare that implementation to the APL version.

We include PyTorch as a reference, but were unable to include a directly comparable reimplementation of PyTorch's functionality in another programming language due to space constraints and difficulty of implementation. The authors have some experience doing such things in various languages, including leveraging "supporting tools" like numpy, to implement even simpler machine learning projects than the U-net architecture described here, but even such simpler implementations with the full support of a system like numpy are far too complex, verbose, and difficult, to say nothing of its performance, for inclusion literally and in full within the body of a conference paper. Even less would be the feasability of including a performant reimplementation of PyTorch's functionality as used in this paper in its totality within the body of a single paper. Nobody would even attempt to do this.

Yet, with APL, we are able to include 2 copies of every expression, a typeset and a source-style formatting version, in the paper, in full detail, and these expressions implement the entire semantics of a neural network from the ground up in complete detail. Even more, when we compared these APL expressions to the mathematical expressions used in various online resources on neural network programming to describe the same things, our APL expressions were actually shorter and structurally simpler than the standard mathematical expressions for the same concepts!

Thus, these single APL expressions are sufficiently expressive, complete, and concise that they can subsume both the mathematical expressions associated with the operations as well as libraries like PyTorch, while also achieving performance that is "shockingly" close to that of highly tuned and labored over domain-specific libraries such as PyTorch, without a single domain-specific optimization anywhere.

These results become even more worthy of note by the fact that other than the use of the APL paradigm and its notation there are zero new novel concepts applied. The compiler itself uses no special or novel optimizations. There are no special implementation details, no novel data structures or other programming language semantics. There is no special code generation. The only novel thing is the use of APL as the language/semantics. There are no domain-specific features optimized for neural networks anywhere in the programming stack.

In a paper of this sort, we cannot hope to provide a full treatment of CNNs, so we have chosen a network of sufficient complexity as well as sufficient history that a reader may find ample discussion of the architecture and possible variations elsewhere, and mimicked its implementation as closely to the standard as possible. Likewise, we cannot hope to turn the reader into an expert at APL within the space of a single paper; the language is too alien for that.

What we can do is provide a demonstration of the power of the notation and inspire the reader to pursue such ideas in more depth. We intend to provide extended resources that would not fit in the paper for understanding these ideas, but these will not fit within the space limitations of the paper.

With all of that said, the reviewers have highlighted important points of clarity that will permit us to more fully and simply explain some of the pain points that tripped up the reviewers, and this should greatly simplify the parsing of some of the expressions, though the intuition of the array programming model will still be foreign to some. A few careful simplifications and removals will eliminate many of the specific points brought up by reviewers and simplify the overall paper, while also giving us more space to dedicate to explicating the expressions in more detail.

...

### 2022-05-05 2022-icfp-unet/Paper/Response.txt

The authors thank the reviewers for their thoughtful responses.

Much like traditional math notation, recursion, static typing, or laziness, APL is cryptic and challenging to the uninitiated. We do not expect uninitiated readers to easily apprehend the APL expressions. We *do* expect readers to perceive the novelty and impact of our results. Ironically, while APL appears complex, PyTorch *is* complex.

Calling the PyTorch API is easier than using APL, but it is not the comparison we make and we will clarify this point in the final paper. The better comparison is our APL implementation vs. PyTorch's underlying code. Our paper contains the complete, from scratch APL code developed almost entirely by a single developer. APL itself has no domain-specifc support for machine learning or neural networks. PyTorch contains over 200,000 lines of code from over 2,000 contributors specifically targeting ML. We omitted the equivalent implementation of PyTorch for space concerns.

Likewise, the APL code is more simple, concise, and well-defined than the equivalent traditional mathematical equations. We do not believe APL to be more cryptic to the uninitiated than the same expressions written in traditional math would be to someone uninitiated in traditional math notation.

APL is the only novel technique in our entire computational stack. The compiler uses no special optimizations and the runtime depends on no specialized machine learning code in order to see the runtime performance that we do.

These results are surprising and remarkable; the reader can grasp these results without understanding the details of the APL expressions. That we did not use any novelty other than APL makes the novelty and impact of our results that much greater. We hope that such results inspire and motivate readers to spend the time and effort to dive more deeply into the APL with the help of external resources that we could not fit into the paper but intend to provide.

As the reviewers note, a paper of this size cannot do everything. It can neither teach the reader to rapidly comprehend APL nor elucidate the nuances of CNNs to those uninitiated in APL. We can, however, make the contribution of the paper and the exposition of our code as clear as space permits. Embracing the reviewers' salutary comments, we propose making the following changes in the final paper to improve the clarity of the paper's contribution and exposition:

### 2022-05-05 2022-icfp-unet/Paper/Response.txt

Much like traditional math notation, recursion, static typing, or laziness, APL is cryptic and challenging to the uninitiated. We do not expect uninitiated readers to easily apprehend the APL expressions. We *do* expect readers to perceive the novelty and impact of our results. Ironically, while APL appears complex, PyTorch *is* complex.

Calling the PyTorch API is easier than using APL, but it is not the comparison we make and we will clarify this point in the final paper. The better comparison is our APL implementation vs. PyTorch's underlying code. Our paper contains the complete, from scratch APL code developed almost entirely by a single developer. APL itself has no domain-specifc support for machine learning or neural networks. PyTorch contains over 200,000 lines of code from over 2,000 contributors specifically targeting ML. We omitted the equivalent implementation of PyTorch for space concerns.

Likewise, the APL code is more simple, concise, and well-defined than the equivalent traditional mathematical equations. We do not believe APL to be more cryptic to the uninitiated than the same expressions written in traditional math would be to someone uninitiated in traditional math notation.

APL is the sole novelty in our entire stack. The compiler uses no special optimizations and the runtime depends on no specialized machine learning code in order to see the runtime performance that we do. Our key result is as much about what we didn't do as what we did.

@@

These results are surprising and remarkable; the reader can grasp these results without understanding the details of the APL expressions. That we did not use any novelty other than APL makes the novelty and impact of our results that much greater. We hope that such results inspire and motivate readers to spend the time and effort to dive more deeply into the APL with the help of external resources that we could not fit into the paper but intend to provide.

As the reviewers note, a paper of this size cannot do everything. It can neither teach the reader to rapidly comprehend APL nor elucidate the nuances of CNNs to those uninitiated in APL. We can, however, make the contribution of the paper and the exposition of our code as clear as space permits. Embracing the reviewers' salutary comments, we propose making the following changes in the final paper to improve the clarity of the paper's contribution and exposition:

## PROPOSED CHANGES

1. Remove all sections and discussions related to different implementations of the stencil function and operator. This is unnecessary to understand the main point of the paper and it is less accessible or interesting to a general audience.

2. Simplify the discussion of Rank since we do not use Rank in its full generality.

3. Include fuller descriptions of the APL primitives used include those mentioned by the reviewers such as the turnstile ⊢.

4. We will include type signatures for APL primitives to help assist in building an intuition of the APL code.

5. Discuss our verification of the correctness the APL version, which includes verifying the output of the APL code against the output of the PyTorch implementation.

6. Provide references to external resources that provide a more expanded treatment of the APL code.

7. Clarify the introduction and discussion to focus specifically on our key results and remove discussion around tangential topics.

8. Add an appendix with our APL code transliterated into a Python-esque pseudocode as an aid to parsing the expressions

9. Make it clear that our primary contribution is not comparing the usability of PyTorch and its performance against APL.

10. Make it clear that the compiler is not doing any special optimizations behind the scenes.

## ADDITIONAL, SPECIFIC REVIEWER COMMENTS RESPONSES

Additional comments on direct points made by reviewers:

* Our back-propagation code was written manually, without auto-differentiation.

* We avoided deviating from the U-net architecture and training practices found in the origional paper in order to maximize the applicability of external resources in understanding our architecture and to avoid introducing any novel CNN strategies that might conflate the point of the paper.

* The use of bilinear upsampling would not materially alter the design and appearance of the transposed convolutional layers except to have a little more mathematical operations. In APL at least, they don't represent a fundamentally different computational flavor.

* The rules and code are duplicates of one another with the rules describing the pure computation of the layer modulo any threading of values in and out of a layer-store. The source code given is the same expression but with the code that gets and sets the stored layers in the network included within the code

* Re: Debugging. Dyalog APL includes a complete line-oriented debugger with traps, stops, restarting, stack inspection, frame-localized REPLs, and so forth. There are also specific display and graphics utilities for rendering the layers as bitmaps or graphs for tracking their progress during the debugging of the training process

* We chose not to optimize the APL code because we were primarily interested in establishing a baseline of performance to see how the most direct and naive implemenations of the neural network in APL would look and perform without doing any sophisticated work to alter that behavior.

* Re: line 1019, this is part of the sections that we propose deleting in their entirety from the paper

* re: line 1104, We will clarify that we are not trying to equate the APL code to calling PyTorch's API, but rather, comparing APL against what the underlying implementation of PyTorch looks like and its complexity.

* "What is needed is more detail on the algorithm, for non-neural-network
wizards. Then a *much* more detailed development of how the
convolutions and other computational elements of the application are
implemented in APL." 
  --- We will include references to detailed expositions of the U-net architecture, since a detailed explanation of the algorithm if one is not familiar with neural network programming would not fit within the paper. Likewise, we will reference a much more detailed development of the APL code, too. Neither of these are appropriate within the paper itself, at the very least because of the space constraints.

* "My dissatisfaction with this paper is that the heart of the paper,
which is the development of the CNN application in APL, is pretty
incomprehensible. APL is a tough language to explicate with clarity."
  --- The final paper will make the results and impact more accessible to those who do not fully grasp the APL. A full understanding of the APL involved is not something that the reader should expect to be fast or easy, and we recognize that it can take significant time to do so, more than would be expected of the time to read a paper in which one were more familiar with the languages and ideas. We think there is an inherent element of difficulty in this problem because of the nature of the language and programming paradigm that we cannot escape: motivated readers who wish to learn the APL will have to accept that doing so will take a longer time than they might first expect. We believe that the reader can still benefit without this heavy investment if they choose not to dive deeply into the APL.
